---
title: "Project Part 2"
author: "Maryam Imran"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE)
```

**Repository URL:** https://github.com/maryamimran12/stat3080-project

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load required packages
library(tidyverse)

# Read in the unemployment data (wide format: one column per year)
unemp <- read_csv("state_unemployment.csv")

# Pivot to long format and convert Year to integer
unemp_long <- unemp %>%
  pivot_longer(cols = `2019`:`2024`, names_to = "Year", values_to = "Rate") %>%
  mutate(Year = as.integer(Year))

# Filter to 2024 only for this analysis
unemp_2024 <- unemp_long %>%
  filter(Year == 2024)
```

## Introduction

This report continues the analysis from Part 1, which explored how annual unemployment
rates vary across U.S. states and Census regions. The exploratory analysis revealed that
in 2024, the West had the highest mean unemployment rate (4.02%) and the Midwest the
lowest (3.46%), with notable variation in spread across regions. This report formally tests
whether those regional differences are statistically significant.

**Research question:** Do mean annual unemployment rates differ significantly across U.S.
Census regions in 2024?

## Methods

A one-way analysis of variance (ANOVA) is used to test whether at least one Census
region has a different mean unemployment rate from the others in 2024. ANOVA is
appropriate here because the response variable (state unemployment rate) is continuous,
there are four independent groups (Northeast, South, Midwest, West), and the goal is to
compare means across more than two groups using a single standalone procedure.

The hypotheses are:

$$H_0: \mu_{\text{NE}} = \mu_{\text{S}} = \mu_{\text{MW}} = \mu_{\text{W}}$$
$$H_a: \text{At least one regional mean unemployment rate differs from the others}$$

A significance level of $\alpha = 0.05$ is used.

**Assumption verification.** One-way ANOVA requires (1) independence of observations,
(2) approximate normality within each group, and (3) equal variances across groups.

*Independence:* Each observation is a U.S. state or D.C. States are distinct geographic
and economic units; their unemployment rates are collected independently by the BLS LAUS
program. This assumption is reasonably met.

*Normality:* With group sizes ranging from 9 (Northeast) to 17 (South), the Central Limit
Theorem cannot be invoked. Normal Q-Q plots for each region are examined instead
(Figure 1). All four plots show points tracking closely along the reference line with no
severe departures, supporting approximate normality within each region.

*Equal variances:* The standard deviations from Part 1 (Midwest: 0.96, Northeast: 0.79,
West: 0.84, South: 0.68) are all within a factor of 1.5 of one another. The residual plot
(Figure 2) shows no systematic fanning or clustering across groups. The equal variance
assumption is reasonably satisfied.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.5, fig.width=6.5, fig.cap="Normal Q-Q plots of 2024 state unemployment rates by Census region. Points track closely along the reference line in all four regions, supporting approximate normality."}
# Q-Q plots by region to assess normality assumption
ggplot(unemp_2024, aes(sample = Rate)) +
  stat_qq(color = "#4E79A7", size = 1.8) +
  stat_qq_line(color = "gray40", linewidth = 0.7) +
  facet_wrap(~Region, nrow = 2) +
  labs(
    title = "Figure 1: Normal Q-Q Plots by Census Region (2024)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles (Unemployment Rate %)"
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10),
        strip.text = element_text(face = "bold"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=2.5, fig.width=6, fig.cap="Residuals from the one-way ANOVA model plotted against fitted values (regional means). The spread of residuals is similar across all four regions, supporting the equal variance assumption."}
# Fit the ANOVA model to extract residuals for assumption checking
fit_check <- aov(Rate ~ Region, data = unemp_2024)

# Residual plot to assess equal variance assumption
data.frame(
  Fitted = fitted(fit_check),
  Residuals = residuals(fit_check),
  Region = unemp_2024$Region
) %>%
  ggplot(aes(x = Fitted, y = Residuals, color = Region)) +
  geom_point(size = 2.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Figure 2: Residuals vs. Fitted Values",
    x = "Fitted Values (Regional Mean Unemployment Rate %)",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 10))
```

\newpage

## Results

```{r}
# One-way ANOVA: test whether mean unemployment rate differs across Census regions
# Response: 2024 state unemployment rate (continuous)
# Factor: Census region (4 levels: Midwest, Northeast, South, West)
anova_result <- aov(Rate ~ Region, data = unemp_2024)

# Display the ANOVA table
summary(anova_result)
```

The ANOVA table shows an F-statistic of `r round(summary(anova_result)[[1]]$"F value"[1], 2)`
on 3 and `r summary(anova_result)[[1]]$Df[2]` degrees of freedom, with a p-value of
`r round(summary(anova_result)[[1]]$"Pr(>F)"[1], 4)`. Since this p-value
`r ifelse(summary(anova_result)[[1]]$"Pr(>F)"[1] < 0.05, "is less than", "exceeds")`
$\alpha = 0.05$, the result is
`r ifelse(summary(anova_result)[[1]]$"Pr(>F)"[1] < 0.05, "statistically significant", "not statistically significant")`.
We `r ifelse(summary(anova_result)[[1]]$"Pr(>F)"[1] < 0.05, "reject $H_0$ and conclude that at least one Census region has a mean unemployment rate that differs from the others in 2024.", "fail to reject $H_0$; the data do not provide sufficient evidence that mean unemployment rates differ across Census regions in 2024.")`

## Conclusions

The one-way ANOVA provides
`r ifelse(summary(anova_result)[[1]]$"Pr(>F)"[1] < 0.05, "statistically significant evidence", "insufficient evidence")`
that mean state unemployment rates differ across U.S. Census regions in 2024
(F = `r round(summary(anova_result)[[1]]$"F value"[1], 2)`,
p = `r round(summary(anova_result)[[1]]$"Pr(>F)"[1], 4)`).
In practical terms, while the exploratory analysis in Part 1 revealed descriptive differences
in regional means — the West averaging 4.02% and the Midwest 3.46% — the within-region
variability is large enough that these differences are not statistically distinguishable at the
0.05 level. Census region alone does not appear to be a statistically significant predictor
of state unemployment rates in 2024.

**Limitations.** First, the ANOVA identifies that at least one regional mean differs but
does not specify which pairs differ; a post-hoc procedure (e.g., Tukey HSD) would be
needed for pairwise comparisons. Second, the analysis treats D.C. as part of the South,
which may inflate that region's mean given D.C.'s uniquely concentrated federal labor
market. Third, unemployment rates are model-based estimates subject to periodic BLS
revision, introducing a small degree of measurement uncertainty.

**Generalizability.** Because the BLS LAUS data cover all 50 states and D.C. — not a
sample — the results describe the actual 2024 population of U.S. state-level labor markets.
Caution is warranted in generalizing across years, as regional patterns shift with economic
cycles, policy changes, and external shocks such as the COVID-19 pandemic.

\newpage

## References

Bureau of Labor Statistics, U.S. Department of Labor. (2021). *Regional and State
Unemployment -- 2020 Annual Averages* (USDL-21-0363). Retrieved from
https://www.bls.gov/news.release/archives/srgune_03032021.pdf

Bureau of Labor Statistics, U.S. Department of Labor. (2023). *Regional and State
Unemployment -- 2022 Annual Averages* (USDL-23-0397). Retrieved from
https://www.bls.gov/news.release/archives/srgune_03012023.pdf

Bureau of Labor Statistics, U.S. Department of Labor. (2025). *Regional and State
Unemployment -- 2024 Annual Averages* (USDL-25-0294). Retrieved from
https://www.bls.gov/news.release/pdf/srgune.pdf

U.S. Census Bureau. (2024). *Census Regions and Divisions of the United States*.
Retrieved from https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf

AI Attribution: Portions of this report including the R Markdown structure, code, and
written prose were drafted with assistance from Claude (Anthropic, claude.ai). All data
values were verified against official BLS publications. All output was reviewed and edited
by the author in accordance with UVA academic integrity guidelines.
