---
title: "Printed tables"
output: pdf_document
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, highlight=FALSE, message=FALSE)
```

```{r, warning=FALSE}
library(tidyverse)
library(dslabs)
library(janitor)
library(knitr)

temp_carbon2 <- temp_carbon %>%
  mutate(decade = year - year %% 10) %>%
  filter(decade >= 1900 & decade <= 2000)

table1 <- temp_carbon2 %>%
  group_by(decade) %>%
  summarize(minco2 = min(carbon_emissions, na.rm = TRUE),
            medco2 = median(carbon_emissions, na.rm = TRUE),
            meanco2 = mean(carbon_emissions, na.rm = TRUE),
            sdco2 = sd(carbon_emissions, na.rm = TRUE),
            maxco2 = max(carbon_emissions, na.rm = TRUE)) 

table2 <- temp_carbon2 %>%
  group_by(decade) %>%
  summarize(minco2 = min(carbon_emissions, na.rm = TRUE),
            medco2 = median(carbon_emissions, na.rm = TRUE),
            meanco2 = mean(carbon_emissions, na.rm = TRUE),
            sdco2 = sd(carbon_emissions, na.rm = TRUE),
            maxco2 = max(carbon_emissions, na.rm = TRUE),
            mintempa = min(temp_anomaly, na.rm = TRUE),
            medtempa = median(temp_anomaly, na.rm = TRUE),
            meantempa = mean(temp_anomaly, na.rm = TRUE),
            sdtempa = sd(temp_anomaly, na.rm = TRUE),
            maxtempa = max(temp_anomaly, na.rm = TRUE)) 

table3 <- data.frame(variable.name = c("Year", 
                                       "Global temperature anomaly",
                                       "Land temperature anomaly",
                                       "Ocean temperature anomaly",
                                       "Carbon emissions"),
                    description = c("Year the measurement was taken", 
                                    "Global annual mean temperature anomaly relative to the 20th century mean temperature in Celsius",
                                    "Annual mean temperature anomaly on land relative to the 20th century mean temperature in Celsius",
                                    "Annual mean temperature anomaly over oceans relative to the 20th century mean temperature in Celsius",
                                    "Annual carbon emissions in millions of metric tons"))
```


## Printed tibble (numeric)
```{r}
table1
```


\vspace{24pt}

## Printed kable (numeric)
```{r}
table1 %>%
  kable()
```


\vspace{24pt}

## Printed kable with variable names and caption (numeric)
```{r}
table1 %>%
  kable(col.names = c("Decade", "Minimum", "Median", "Mean", "StDev", "Maximum"),
        caption = "Annual carbon emissions")
```


\vspace{24pt}

## Printed kable with formatting (numeric)
```{r}
table1 %>%
  kable(col.names = c("Decade", "Minimum", "Median", "Mean", "StDev", "Maximum"),
        caption = "Annual carbon emissions",
        digits = 1,
        align = "cccccc")
```


\newpage

## Printed tibble (numeric)
```{r}
table2 
```


\vspace{24pt}

## Printed kable (numeric)
```{r}
table2 %>%
  kable()
```


\newpage

## Printed kable with variable names and caption (numeric)
```{r}
table2 %>%
  kable(col.names = c("Decade", "Minimum carbon", "Median carbon", "Mean carbon", "StDev carbon", "Maximum carbon", 
                      "Minimum anomaly", "Median anomaly", "Mean anomaly", "StDev anomaly", "Maximum anomaly"),
        caption = "Annual carbon emissions and global temperature anomalies")
```


\vspace{24pt}

## Printed kable with formatting (numeric)
```{r}
table2 %>%
  kable(col.names = c("Decade", "Minimum carbon", "Median carbon", "Mean carbon", "StDev carbon", "Maximum carbon", 
                      "Minimum anomaly", "Median anomaly", "Mean anomaly", "StDev anomaly", "Maximum anomaly"),
        caption = "Annual carbon emissions and global temperature anomalies",
        digits = 1,
        align = "ccccccccccc")
```


\vspace{24pt}

## Printed kable with formatting (numeric)
```{r}
table2 %>%
  select(decade, minco2:maxco2) %>%
  kable(col.names = c("Decade", "Minimum carbon", "Median carbon", "Mean carbon", "StDev carbon", "Maximum carbon"),
        caption = "Annual carbon emissions and global temperature anomalies",
        digits = 1,
        align = "ccccccccccc")

table2 %>%
  select(decade, mintempa:maxtempa) %>%
  kable(col.names = c("Decade", "Minimum anomaly", "Median anomaly", "Mean anomaly", "StDev anomaly", "Maximum anomaly"),
        digits = 1,
        align = "ccccccccccc")
```





\newpage

## Printed data frame (character)
```{r}
table3
```


\vspace{24pt}

## Printed kable with variable names and caption (character)
```{r}
table3 %>%
  kable(col.names = c("Variable name", "Variable description"),
        caption = "Data dictionary")
```


\newpage

## Printed kable with kableExtra formatting (character)
```{r}
library(kableExtra)

table3 %>%
  kable(col.names = c("Variable name", "Variable description"),
        caption = "Data dictionary")
```


\newpage

## Printed kable with kableExtra formatting (character)
```{r}
table3 %>%
  kable(col.names = c("Variable name", "Variable description"),
        caption = "Data dictionary",
        booktabs = TRUE,
        linesep = "\\addlinespace") %>%
  kable_styling(latex_options = "hold_position", 
                full_width=TRUE) %>%
  column_spec(1, width="1.5in") 
```


\vspace{24pt}

## Printed kable with kableExtra formatting (numeric)
```{r}
table2 %>%
  select(decade, meanco2:sdco2, meantempa:sdtempa) %>%
  kable(col.names = c("Decade", "Mean", "StDev", "Mean", "StDev"),
        caption = "Annual carbon emissions and global temperature anomalies",
        digits = 1,
        align = "ccccc",
        booktabs = TRUE,
        linesep = "") %>%
  kable_styling(latex_options = "hold_position") %>%
  add_header_above(c(" " = 1, "Carbon" = 2, "Temp anomaly" = 2))
```


