---
title: "Project Part 1"
fontsize: 12pt
geometry: margin=1in
output:
  pdf_document: default
  word_document: default
urlcolor: black
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---
**Maryam Imran**  


**Repository URL:** https://github.com/UVA-STAT3080/student-Sp26
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)

unemp <- read_csv("state_unemployment.csv")

unemp_long <- unemp %>%
  pivot_longer(cols = `2019`:`2024`,
               names_to = "Year",
               values_to = "Rate") %>%
  mutate(Year = as.integer(Year))
```


## Introduction

The unemployment rate is one of the most closely watched indicators of 
economic health in the United States. While national figures dominate 
headlines, state-level rates can vary substantially, reflecting differences 
in industry composition, demographics, and regional economic conditions. 
This report investigates the following research question: **How do annual 
unemployment rates vary across U.S. states and Census regions, and how have 
these rates changed from 2019 to 2024?**

This time window spans the pre-pandemic economy (2019), the COVID-19 labor 
market shock (2020), and the subsequent recovery period (2021--2024). 
Understanding geographic patterns in unemployment can inform economic policy 
and highlight which regions are most vulnerable to major labor market 
disruptions.


## Data Summary

The data come from the Bureau of Labor Statistics (BLS) Local Area 
Unemployment Statistics (LAUS) program, a federal-state cooperative effort 
administered by the U.S. Department of Labor. The program produces annual 
estimates of unemployment for all 50 U.S. states and the District of 
Columbia, derived by combining data from the Current Population Survey (CPS), 
the Current Employment Statistics (CES) survey, and state unemployment 
insurance (UI) systems. Because estimates are produced for every state and 
D.C., the data represent the entire target population of U.S. state-level 
jurisdictions, not a sample.

Annual average unemployment rates for 2019 through 2024 were compiled from 
four official BLS annual averages news releases (see References). A Region 
variable was added using the U.S. Census Bureau's standard four-region 
classification (Northeast, South, Midwest, West).

Two limitations are worth noting. First, LAUS estimates are periodically 
revised as benchmark data become available. Second, the unemployment rate 
excludes discouraged workers who have stopped searching for employment, which 
may understate true labor market weakness during severe downturns. Despite 
these limitations, the data are appropriate for the research question because 
they provide consistent, official government estimates across three 
economically distinct periods.
\newpage


## Exploratory Analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

# Numerical Summary 1: Regional statistics for 2024
region_summary <- unemp %>%
  group_by(Region) %>%
  summarise(
    N      = n(),
    Mean   = round(mean(`2024`, na.rm = TRUE), 2),
    Median = round(median(`2024`, na.rm = TRUE), 2),
    SD     = round(sd(`2024`, na.rm = TRUE), 2),
    Min    = round(min(`2024`, na.rm = TRUE), 2),
    Max    = round(max(`2024`, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean))

kable(region_summary,
      caption = "\\textbf{Summary Statistics of State Unemployment Rates (\\%) by Census Region, 2024}",
      col.names = c("Region", "N", "Mean", "Median", "Std Dev", "Min", "Max"),
      align = "lrrrrrr",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE,
                font_size = 13) %>%
  column_spec(1, width = "2.5cm") %>%
  column_spec(2:7, width = "1.5cm")
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

# Numerical Summary 2: Five highest and five lowest states in 2024
top5_high <- unemp %>%
  arrange(desc(`2024`)) %>%
  slice(1:5) %>%
  select(State, Region, `2024`) %>%
  mutate(Group = "Highest 5")

top5_low <- unemp %>%
  arrange(`2024`) %>%
  slice(1:5) %>%
  select(State, Region, `2024`) %>%
  mutate(Group = "Lowest 5")

kable(bind_rows(top5_high, top5_low),
      caption = "\\textbf{Five Highest and Five Lowest State Unemployment Rates, 2024}",
      col.names = c("State", "Census Region", "Rate (%)", "Group"),
      align = "llrl",
      booktabs = TRUE,
      longtable = FALSE) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE,
                font_size = 13) %>%
  column_spec(1, width = "3.5cm") %>%
  column_spec(2, width = "3cm") %>%
  column_spec(3:4, width = "2cm")
```
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7, fig.pos="H", fig.cap="Distribution of state unemployment rates by Census region in 2019 (pre-pandemic) and 2024. Each box shows the median, interquartile range, and spread of state rates within the region. Source: BLS LAUS Program (2025)."}
# Graphical Summary 1: Side by side boxplots 2019 vs 2024
unemp_long %>%
  filter(Year %in% c(2019, 2024)) %>%
  mutate(Year = factor(Year)) %>%
  ggplot(aes(x = Region, y = Rate, fill = Year)) +
  geom_boxplot(width = 0.5, outlier.shape = 21,
               outlier.size = 2, alpha = 0.85) +
  scale_fill_manual(values = c("2019" = "#B07AA1",
                               "2024" = "#4E79A7")) +
  labs(
    title    = "Unemployment Rates Rose Across All U.S. Regions from 2019 to 2024",
    subtitle = "Comparing pre-pandemic and current state-level labor market conditions",
    x        = "Census Region",
    y        = "Unemployment Rate (%)",
    fill     = "Year",
    caption  = "Source: U.S. Bureau of Labor Statistics, LAUS Program (2025)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7, fig.pos="H", fig.cap="Average annual state unemployment rate by Census region from 2019 to 2024. Each panel shows one region's trend. The sharp 2020 spike reflects the COVID-19 pandemic. Source: BLS LAUS Program (2025)."}
# Graphical Summary 2: Faceted line chart by region 2019-2024
unemp_long %>%
  group_by(Region, Year) %>%
  summarise(Mean_Rate = mean(Rate, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Year, y = Mean_Rate)) +
  geom_line(linewidth = 1.3, color = "#4E79A7") +
  geom_point(size = 3.5, color = "#4E79A7") +
  geom_area(alpha = 0.15, fill = "#4E79A7") +
  facet_wrap(~Region, nrow = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_y_continuous(limits = c(2, 11),
                     breaks = seq(2, 11, by = 2)) +
  labs(
    title   = "COVID-19 Caused a Sharp Spike in Unemployment Across All Regions",
    subtitle = "Average state unemployment rate by Census region, 2019-2024",
    x       = "Year",
    y       = "Avg. Unemployment Rate (%)",
    caption = "Source: U.S. Bureau of Labor Statistics, LAUS Program (2025)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"),
        strip.text = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
\newpage


## Conclusions

The exploratory analysis reveals several notable patterns in U.S. state-level
unemployment rates across the 2019--2024 period.

**Regional variation.** Table 1 shows that the West had the highest mean
unemployment rate in 2024 at 4.02%, driven largely by California and Nevada,
whose large labor forces are concentrated in industries sensitive to economic
cycles. The Midwest had the lowest mean rate at 3.46% and the smallest
standard deviation, indicating both lower unemployment and greater consistency
across its states.

**State-level extremes.** Table 2 confirms that the five highest-unemployment
states in 2024 were Nevada (5.6%), California (5.3%), District of Columbia
(5.2%), Kentucky (5.1%), and Illinois (5.0%). The five lowest were South Dakota
(1.8%), Vermont (2.3%), North Dakota (2.4%), New Hampshire (2.6%), and Nebraska
(2.8%), predominantly small Midwest and Northeast states with tight labor markets.

**COVID-19 shock and recovery.** Figure 2 shows a dramatic spike in 2020 across
all four regions. The West experienced the sharpest increase, driven by Hawaii
and Nevada whose tourism-dependent economies were hit hardest by pandemic
shutdowns. All regions recovered close to pre-pandemic levels by 2022, though
the West remained the highest through 2024.

**Implications for Part 2.** These patterns suggest that Census region explains
a meaningful portion of state-level unemployment variation. A formal test of
whether mean unemployment rates differ significantly across regions will be
pursued in Part 2.





\newpage

## References

Bureau of Labor Statistics, U.S. Department of Labor. (2021). *Regional and 
State Unemployment -- 2020 Annual Averages* (USDL-21-0363). Retrieved from
https://www.bls.gov/news.release/archives/srgune_03032021.pdf

Bureau of Labor Statistics, U.S. Department of Labor. (2023). *Regional and 
State Unemployment -- 2022 Annual Averages* (USDL-23-0397). Retrieved from
https://www.bls.gov/news.release/archives/srgune_03012023.pdf

Bureau of Labor Statistics, U.S. Department of Labor. (2025). *Regional and 
State Unemployment -- 2024 Annual Averages* (USDL-25-0294). Retrieved from
https://www.bls.gov/news.release/pdf/srgune.pdf

U.S. Census Bureau. (2024). *Census Regions and Divisions of the United 
States.* Retrieved from
https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf

*AI Attribution: Portions of this report including the R Markdown structure, 
code, and written prose were drafted with assistance from Claude (Anthropic, 
claude.ai). All data values were verified against official BLS publications. 
All output was reviewed and edited by the author in accordance with UVA 
academic integrity guidelines.*
